---
title: "Session 8: Binary Dependent Variables"
author: "Patrick Shea"
format: html
execute:
  warning: false
  message: false
---

```{r}
#| label: setup

## Set your working directory to the ICPSR_regression folder
## Change the path below to match your computer
# setwd("~/ICPSR_regression")
setwd("C:/Users/Patrick Shea/Dropbox/My Projects/Spring 2026/ICPSR/ICPSR_OLS26")

rm(list = ls())

# install.packages(c("faraway", "foreign", "lattice", "stargazer", "mfx",
#                     "effects", "reshape2", "ResourceSelection", "ROCR",
#                     "pROC", "heatmapFit", "separationplot", "interplot"))
library(faraway)
require(foreign)
```

---

# Part 1: The Linear Probability Model and Its Limits

## Supreme Court decisions data

Our first example uses data on U.S. Supreme Court habeas corpus decisions. The outcome variable `propetit` equals 1 if the Court ruled in favor of the petitioner and 0 otherwise. Our key predictor is the Court's ideological orientation (`liberal`), scaled from 0 to 1.

```{r}
#| label: load-court

mydata <- read.dta("Court.dta")

# Rescale liberalism to a 0--100 scale for easier interpretation
mydata$liberal2 <- round(mydata$liberal * 100, 2)
attach(mydata)
```

```{r}
#| label: explore-court

head(mydata)
summary(mydata)
```

Let's look at the distribution of our key variables:

```{r}
#| label: explore-plots

hist(mydata$liberal2,
     main = "Distribution of Court Liberalism",
     xlab = "Liberalism (0-100 scale)")

library(lattice)
densityplot(mydata$liberal2)

barplot(sort(table(mydata$propetit), decreasing = TRUE), las = 2,
        main = "Distribution of Petition Outcomes")
```

### Why not just use OLS?

Let's plot the binary outcome against liberalism and fit a linear regression:

```{r}
#| label: lpm

plot(propetit ~ liberal2, mydata, xlim = c(0, 100), ylim = c(0, 1),
     xlab = "Liberalism", ylab = "Observed Petition Outcome")

lmod <- lm(propetit ~ liberal2, mydata)
abline(lmod)

sumary(lmod)
```

This is the **linear probability model** (LPM). The line goes straight through the data, but notice the problem: for very low or very high values of liberalism, the predicted probability can go below 0 or above 1. That's nonsensical for a probability. We need a model that constrains predictions to the [0, 1] interval.

---

# Part 2: Logit and Probit Models

Both **logit** and **probit** solve the LPM's problem by passing the linear predictor ($X\beta$) through a nonlinear function that maps any real number onto (0, 1):

- **Logit** uses the logistic function: $P(Y=1) = \frac{e^{X\beta}}{1 + e^{X\beta}}$
- **Probit** uses the cumulative normal: $P(Y=1) = \Phi(X\beta)$

```{r}
#| label: logit-probit

logitmod  <- glm(propetit ~ liberal2, family = binomial(link = logit), data = mydata)
probitmod <- glm(propetit ~ liberal2, family = binomial(link = probit), data = mydata)

summary(logitmod)
summary(probitmod)
```

Comparing all three models side by side:

```{r}
#| label: three-models
#| results: asis

library(stargazer)
stargazer(lmod, logitmod, probitmod, type = "text")
```

The coefficients across logit and probit are on different scales (log-odds vs. z-scores), so they're not directly comparable in magnitude. What matters is the sign, significance, and --- most importantly --- the **predicted probabilities**.

---

# Part 3: Interpreting Coefficients

## Predicted probabilities at specific values

The raw logit coefficient tells us the change in **log-odds** per unit increase in liberalism. To make this concrete, we compute predicted probabilities at the minimum, median, and maximum of liberalism:

```{r}
#| label: pred-probs-specific

summary(liberal2)

# At min (21), median (41), and max (83.11)
# Logit predictions
ilogit(-3.2828 + 0.06929 * 21)
ilogit(-3.2828 + 0.06929 * 41)
ilogit(-3.2828 + 0.06929 * 83.11)

# Probit predictions (for comparison)
pnorm(-1.9731 + 0.04176 * 21)
pnorm(-1.9731 + 0.04176 * 41)
pnorm(-1.9731 + 0.04176 * 83.11)
```

Notice how logit and probit give nearly identical predicted probabilities, even though the raw coefficients look different.

## Marginal effects

A **marginal effect** answers the question: at the average case, how much does a one-unit increase in liberalism change the *probability* of a pro-petitioner decision?

```{r}
#| label: marginal-effects

library(mfx)
logitmfx(propetit ~ liberal2, data = mydata)
probitmfx(propetit ~ liberal2, data = mydata)
```

Marginal effects are on the probability scale and are directly comparable across logit and probit.

## Odds ratios

For logit models specifically, we can exponentiate the coefficients to get **odds ratios**:

```{r}
#| label: odds-ratios

# Point estimates
exp(logitmod$coefficients)

# With 95% confidence intervals
exp(confint(logitmod))

# Or use the logitor function
logitor(propetit ~ liberal2, data = mydata)
```

An odds ratio of 1.07 means that each additional point of liberalism multiplies the odds of a pro-petitioner decision by 1.07 (a 7% increase in the odds).

---

# Part 4: Visualizing Predictions

## Fitted values vs. the original data

The `predict()` function can return predictions on two scales: the **link** scale (log-odds for logit, z-scores for probit) and the **response** scale (probabilities).

```{r}
#| label: fitted-values

logit.phat2  <- predict(logitmod)                  # Link scale (log-odds)
probit.phat2 <- predict(probitmod)                  # Link scale (z-scores)
logit.phat3  <- predict(logitmod, type = "response")  # Probability scale
probit.phat3 <- predict(probitmod, type = "response") # Probability scale
```

On the link scale, the relationship is linear by construction:

```{r}
#| label: plot-link

plot(logit.phat2 ~ mydata$liberal2, lty = 1, type = "l",
     xlab = "Liberalism", ylab = "Predicted Log Odds")
plot(probit.phat2 ~ mydata$liberal2, lty = 2, type = "l", col = "blue",
     xlab = "Liberalism", ylab = "Predicted Z-Score")
```

On the probability scale, we see the characteristic S-curve:

```{r}
#| label: plot-response

plot(logit.phat3 ~ mydata$liberal2, lty = 1, type = "l",
     xlab = "Liberalism", ylab = "Predicted Probability")
plot(probit.phat3 ~ mydata$liberal2, lty = 2, type = "l", col = "blue",
     xlab = "Liberalism", ylab = "Predicted Probability")
```

## Overlaying logit and probit curves on the data

This is perhaps the most informative plot --- the raw data points with both fitted curves:

```{r}
#| label: overlay-curves

plot(propetit ~ liberal2, mydata,
     xlim = c(0, 100), ylim = c(0, 1),
     xlab = "Liberalism", ylab = "Prob of Granting Relief")
x <- seq(0, 100, 1)
lines(x, ilogit(-3.2828 + 0.06929 * x), lty = 1, col = "red", lwd = 2)
lines(x, pnorm(-1.9731 + 0.04176 * x), lty = 2, col = "blue", lwd = 2)
legend("topleft", legend = c("Logit", "Probit"),
       col = c("red", "blue"), lty = c(1, 2), lwd = 2)
```

The two curves are nearly indistinguishable except at the extreme tails.

## Side-by-side comparison

```{r}
#| label: side-by-side

xb1 <- 0.06929 * liberal2
xb2 <- 0.04176 * liberal2

par(mfrow = c(1, 2))
plot(logit.phat3 ~ xb1, lty = 1, type = "lines", xlim = c(1, 6),
     xlab = "Logit Model", ylab = "Predicted Probability")
plot(probit.phat3 ~ xb2, lty = 2, type = "lines", col = "blue",
     xlim = c(0, 4), xlab = "Probit Model", ylab = "Predicted Probability")
par(mfrow = c(1, 1))
```

## The effects package

The `effects` package automates prediction plots and works for any GLM:

```{r}
#| label: effects-simple

library(effects)
plot(allEffects(logitmod))
```

---

# Part 5: From Log-Odds to Probabilities Step by Step

This section walks through the math of transforming predictions, which is useful for building intuition about what the logit model is actually doing.

```{r}
#| label: transform-logodds

# Start with log-odds predictions
plot(logit.phat2 ~ mydata$liberal2, lty = 1, type = "l",
     xlab = "Liberalism", ylab = "Predicted Log Odds",
     main = "Step 1: Log-Odds (hard to interpret)")
```

Exponentiate to get **odds** ($e^{\text{log-odds}}$):

```{r}
#| label: transform-odds

plot(exp(logit.phat2) ~ mydata$liberal2, lty = 2, type = "l",
     xlab = "Liberalism", ylab = "Predicted Odds",
     main = "Step 2: Odds (better, but still unintuitive)")
```

Convert odds to **probabilities** using $p = \frac{\text{odds}}{1 + \text{odds}}$:

```{r}
#| label: transform-probs

plot((exp(logit.phat2) / (1 + exp(logit.phat2))) ~ mydata$liberal2,
     lty = 2, type = "l",
     xlab = "Liberalism", ylab = "Predicted Probabilities",
     main = "Step 3: Probabilities (most intuitive)")
```

This is identical to calling `predict(logitmod, type = "response")`:

```{r}
#| label: transform-verify

plot(logit.phat3 ~ mydata$liberal2, lty = 2, type = "l",
     xlab = "Liberalism", ylab = "Predicted Probabilities (response)",
     main = "Verification: same curve")
```

## Residuals for binary models

OLS-style residual plots don't work well for binary outcomes, but we can still examine them. **Response residuals** are simply $y - \hat{p}$; **Pearson residuals** standardize by the variance:

```{r}
#| label: residuals

resid.res <- residuals(logitmod, type = "response")
resid.mod <- residuals(logitmod, type = "pear")

plot(resid.res ~ mydata$liberal2, main = "Response Residuals")
plot(resid.mod ~ mydata$liberal2, main = "Pearson Residuals")
```

The banding pattern (two horizontal strips) is normal for binary outcomes --- it's inherent in the data, not a sign of model failure.

---

# Part 6: Multivariate Models

## Adding controls

Just as with OLS, we can include additional predictors. The `update()` function is a shortcut for adding variables to an existing model:

```{r}
#| label: multivariate

logitmod2 <- update(logitmod, . ~ . + usparty + ineffcou + multpet)
summary(logitmod2)
```

```{r}
#| label: multi-compare
#| results: asis

stargazer(logitmod, logitmod2, type = "text")
```

Odds ratios with confidence intervals for the full model:

```{r}
#| label: multi-or

round(exp(cbind(Estimate = coef(logitmod2), confint(logitmod2))), 2)
```

Visualizing effects of each predictor:

```{r}
#| label: multi-effects

plot(allEffects(logitmod))
plot(effect("liberal2", logitmod))
plot(allEffects(logitmod2))
plot(effect("liberal2", logitmod2))
```

## Predictions on different scales

```{r}
#| label: multi-predictions

# Link scale (log-odds)
logit.phat2  <- predict(logitmod)
probit.phat2 <- predict(probitmod)

# Response scale (probabilities)
logit.phat3  <- predict(logitmod, type = "response")
probit.phat3 <- predict(probitmod, type = "response")

# Log-odds scale
plot(logit.phat2 ~ mydata$liberal2, lty = 1, type = "l",
     xlab = "Liberalism", ylab = "Predicted Log Odds")
plot(probit.phat2 ~ mydata$liberal2, lty = 2, type = "l", col = "blue",
     xlab = "Liberalism", ylab = "Predicted Log Odds")

# Probability scale
plot(logit.phat3 ~ mydata$liberal2, lty = 1, type = "l",
     xlab = "Liberalism", ylab = "Predicted Probability")
plot(probit.phat3 ~ mydata$liberal2, lty = 2, type = "l", col = "blue",
     xlab = "Liberalism", ylab = "Predicted Probability")
```

Comparing link-scale predictions vs. response-scale predictions:

```{r}
#| label: scale-comparison

# Untranslated linear predictors
xb1 <- -3.2828 + 0.06929 * liberal2
xb2 <- -1.9731 + 0.04176 * liberal2

# Translated through link function
xb3 <- ilogit(-3.2828 + 0.06929 * liberal2)
xb4 <- pnorm(-1.9731 + 0.04176 * liberal2)

# Probabilities as a function of the linear predictor (S-curve)
plot(logit.phat3 ~ xb1, lty = 1, type = "l",
     xlab = "Linear Predictor (Xb)", ylab = "Predicted Probability")
plot(probit.phat3 ~ xb2, lty = 2, type = "l", col = "blue",
     xlab = "Linear Predictor (Xb)", ylab = "Predicted Probability")

# Probabilities as a function of the translated predictor (should be 45-degree line)
plot(logit.phat3 ~ xb3, lty = 1, type = "l",
     xlab = "Transformed Xb", ylab = "Predicted Probability")
plot(probit.phat3 ~ xb4, lty = 2, type = "l", col = "blue",
     xlab = "Transformed Xb", ylab = "Predicted Probability")
```

---

# Part 7: Another Example --- Cars and Fuel Efficiency

To reinforce these ideas, let's try a completely different application using the built-in `mtcars` dataset. We'll create a binary variable for whether a car has "high" fuel efficiency (above the median MPG):

```{r}
#| label: mtcars-setup

data(mtcars)
mtcars$high_mpg <- ifelse(mtcars$mpg >= median(mtcars$mpg), 1, 0)

model <- glm(high_mpg ~ wt + hp + am,
             family = binomial(link = "logit"), data = mtcars)
```

Now let's generate predicted probabilities across the range of vehicle weight, separately for automatic and manual transmissions:

```{r}
#| label: mtcars-predictions

wt_range   <- seq(min(mtcars$wt), max(mtcars$wt), length.out = 100)
x.auto     <- data.frame(wt = wt_range, hp = mean(mtcars$hp), am = 0)
x.manual   <- data.frame(wt = wt_range, hp = mean(mtcars$hp), am = 1)

prob.auto   <- predict(model, newdata = x.auto, type = "response")
prob.manual <- predict(model, newdata = x.manual, type = "response")

plot_data <- data.frame(Weight = wt_range,
                        Automatic = prob.auto,
                        Manual = prob.manual)

library(reshape2)
plot_data_melt <- melt(plot_data, id.vars = "Weight",
                       variable.name = "Transmission",
                       value.name = "Probability")

library(ggplot2)
ggplot(plot_data_melt, aes(x = Weight, y = Probability, color = Transmission)) +
  geom_line(linewidth = 1.2) +
  xlab("Vehicle Weight (1000 lbs)") +
  ylab("Predicted Probability of High MPG") +
  scale_color_manual(values = c("blue", "red"),
                     labels = c("Automatic", "Manual")) +
  theme_minimal()
```

As weight increases, the probability of high fuel efficiency drops sharply --- and manual transmissions maintain higher probabilities at every weight level.

---

# Part 8: Model Diagnostics

With binary outcomes, $R^2$ and standard residual plots aren't particularly informative. Instead, we use specialized diagnostics. Let's refit our Court models to demonstrate:

```{r}
#| label: refit-court

logitmod  <- glm(propetit ~ liberal2, family = binomial(link = logit), data = mydata)
logitmod2 <- update(logitmod, . ~ . + usparty + ineffcou + multpet)

y     <- mydata$propetit
pred1 <- predict(logitmod, type = "response")
pred2 <- predict(logitmod2, type = "response")
```

## Hosmer-Lemeshow test

This test groups observations into deciles of predicted probability and compares observed vs. expected event counts. A significant result ($p < 0.05$) suggests poor fit:

```{r}
#| label: hosmer-lemeshow

library(ResourceSelection)
hoslem.test(mydata$propetit, fitted(logitmod))
hoslem.test(mydata$propetit, fitted(logitmod2))
```

## ROC curves

A **Receiver Operating Characteristic** curve shows the trade-off between sensitivity (true positive rate) and specificity (1 - false positive rate) across all possible classification thresholds.

Using the `ROCR` package (with color for threshold values):

```{r}
#| label: roc-rocr

library(ROCR)

predRoc1 <- prediction(pred1, mydata$propetit)
perfRoc1 <- performance(predRoc1, "tpr", "fpr")
plot(perfRoc1, colorize = TRUE, main = "ROC Curve --- Bivariate Model")

predRoc2 <- prediction(pred2, mydata$propetit)
perfRoc2 <- performance(predRoc2, "tpr", "fpr")
plot(perfRoc2, colorize = TRUE, main = "ROC Curve --- Multivariate Model")
```

Or using the `pROC` package for a cleaner comparison:

```{r}
#| label: roc-proc

library(pROC)
plot.roc(y, pred1, col = "red", main = "ROC Curves: Model 1 (red) vs. Model 2 (blue)")
plot.roc(y, pred2, col = "blue", add = TRUE)
```

## Area Under the Curve (AUC)

The **AUC** summarizes the ROC curve in a single number. AUC = 0.5 means no better than chance; AUC = 1.0 means perfect discrimination:

```{r}
#| label: auc

auc(mydata$propetit, pred1)
auc(mydata$propetit, pred2)
```

## AIC and likelihood ratio test

The **AIC** penalizes complexity (lower is better). The **likelihood ratio test** formally compares nested models:

```{r}
#| label: aic-lr

extractAIC(logitmod)
extractAIC(logitmod2)

library(lmtest)
lrtest(logitmod, logitmod2)
```

## Heatmap fit

The heatmap fit diagnostic (Esarey and Pierce 2012) compares model predictions against smoothed empirical predictions. The 45-degree line represents perfect fit; any deviation suggests model misspecification. Dark shading indicates statistically significant deviations:

```{r}
#| label: heatmap

library(heatmapFit)
heatmap.fit(y, pred1, reps = 1000, legend = FALSE)
heatmap.fit(y, pred2, reps = 1000, legend = FALSE)
```

## Separation plots

**Separation plots** provide an intuitive visual assessment. Observations are sorted left-to-right by predicted probability. Each vertical line represents one case --- dark lines are actual events (pro-petitioner decisions), light lines are non-events. A well-fitting model pushes all dark lines to the right:

```{r}
#| label: separation

library(separationplot)

separationplot(pred = logitmod$fitted.values, actual = logitmod$y,
               type = "rect", line = TRUE, show.expected = TRUE,
               width = 9, height = 1.2, heading = "Bivariate Model")

separationplot(pred = logitmod2$fitted.values, actual = logitmod2$y,
               type = "rect", line = TRUE, show.expected = TRUE,
               width = 9, height = 1.2, heading = "Multivariate Model")
```

For reference, here's what a **perfectly predicted** model would look like --- all events neatly stacked on the right:

```{r}
#| label: separation-perfect

e <- plogis(seq(-10, 3.02, length.out = 500))
f <- c(rep(0, 382), rep(1, 118))
separationplot(pred = e, actual = f, type = "rect", show.expected = TRUE,
               heading = "Perfect Prediction (Reference)")
```

---

# Part 9: Interactions in Logit Models

Interactions work the same way conceptually as in OLS --- the effect of one variable depends on the level of another. But because the logit link function is nonlinear, interpreting interactions requires extra care. Visualization is essential.

## Liberalism x Party

Does the effect of the Court's liberalism on petitioner outcomes depend on whether the U.S. government is a party to the case?

```{r}
#| label: interaction-court

logitint <- glm(propetit ~ liberal2 * usparty + ineffcou + multpet,
                family = binomial(link = logit), data = mydata)
summary(logitint)
```

The `interplot` package visualizes how the marginal effect of one variable changes across values of the other:

```{r}
#| label: interplot-court

library(interplot)
interplot(m = logitint, var1 = "usparty", var2 = "liberal2", hist = TRUE)
interplot(m = logitint, var1 = "liberal2", var2 = "usparty")
```

## Second example: military disputes

To see interactions in a different context, we'll use a dataset on international conflict (`jcrv10.dta`). The question is whether the effect of the military balance (`ratio`) on conflict outcomes depends on the regime type (`dem`):

```{r}
#| label: load-jcrv10

mydata2 <- read.dta("jcrv10.dta")

logitint2 <- glm(wl3 ~ ratio * dem + init + concap + ally1,
                 family = binomial(link = logit), data = mydata2)
summary(logitint2)
```

Marginal effect of the military balance, conditional on democracy:

```{r}
#| label: interplot-jcrv-me

interplot(m = logitint2, var1 = "ratio", var2 = "dem", hist = TRUE) +
  xlab("Democracy") +
  ylab("Marginal Effect of Military Balance") +
  geom_hline(yintercept = 0, linetype = "dashed")
```

We can also plot **conditional predicted probabilities** instead of marginal effects, showing the full probability curves at low and high values of democracy:

```{r}
#| label: interplot-jcrv-pp

interplot(m = logitint2, var1 = "ratio", var2 = "dem", hist = FALSE,
          predPro = TRUE, var2_vals = c(-10, 10)) +
  ggtitle("Conditional Predicted Probabilities") +
  scale_colour_discrete(guide = guide_legend(title = "Dem"),
                        labels = c("Low", "High")) +
  scale_fill_discrete(guide = guide_legend(title = "Dem"),
                      labels = c("Low", "High")) +
  theme(legend.position = c(.2, .8),
        legend.justification = c(0, .5))
```

---

## Summary

| Topic | Key Takeaway |
|-------|-------------|
| LPM vs. logit/probit | OLS on binary outcomes produces impossible predictions; logit and probit constrain probabilities to [0, 1] |
| Interpretation | Use odds ratios, predicted probabilities, or marginal effects --- not raw coefficients |
| Logit vs. probit | Nearly identical predicted probabilities; choice is largely a matter of convention |
| Diagnostics | Hosmer-Lemeshow for calibration; ROC/AUC for discrimination; separation plots for visual assessment |
| Interactions | Conceptually the same as OLS, but must visualize because the nonlinear link makes intuition harder |
