---
title: "Session 7: Panel Data and Two-Stage Models"
author: "Patrick Shea"
format: html
execute:
  warning: false
  message: false
---

## About this document

This is a **Quarto** document --- it mixes explanatory text with executable R code. You can run each code chunk one at a time in RStudio, or click the **Render** button to produce a polished HTML page with all output embedded. No special setup is required beyond having the packages installed.

```{r}
#| label: setup

## Set your working directory to the ICPSR_regression folder
## Change the path below to match your computer
# setwd("~/ICPSR_regression")
rm(list = ls())
setwd("C:/Users/Patrick Shea/Dropbox/My Projects/Spring 2026/ICPSR/ICPSR_OLS26/data")



# install.packages(c("plm", "foreign", "gplots", "car", "stargazer",
#                     "haven", "mediation", "sampleSelection", "AER"))
library(plm)
library(foreign)
library(gplots)
library(car)
library(stargazer)
```

---

# Part 1: Panel Data Methods

## What is panel data?

Up to now, every regression we've run has used **cross-sectional** data --- one snapshot per unit. Panel data (also called longitudinal data) observes the *same* units repeatedly over time. This gives us a big advantage: we can separate changes *within* a unit from differences *between* units, which helps control for unobserved characteristics that might bias our estimates.

## Loading and visualizing panel structure

We'll start with a simple teaching dataset from Princeton that tracks several countries over time. It loads directly from a URL, so no local file is needed.

```{r}
#| label: load-panel

Panel <- read.dta("http://dss.princeton.edu/training/Panel101.dta")
head(Panel)
```

### Visualizing how units change over time

A key first step with panel data is to *see* the structure. The `coplot()` function shows the trajectory of each country in its own panel:

```{r}
#| label: coplot-lines

coplot(y ~ year | country, type = "l", data = Panel)
```

```{r}
#| label: coplot-points

coplot(y ~ year | country, type = "b", data = Panel)
```

The `scatterplot()` function from the `car` package overlays all countries on a single plot, colored by group:

```{r}
#| label: scatter-panel

scatterplot(y ~ year | country, boxplots = FALSE, smooth = TRUE,
            reg.line = FALSE, data = Panel)
```

### The heterogeneity problem

The whole motivation for panel methods is **heterogeneity** --- some countries are systematically higher or lower on our outcome. Let's visualize this directly.

Heterogeneity across countries (unit effects):

```{r}
#| label: plotmeans-country

plotmeans(y ~ country, main = "Heterogeneity across countries", data = Panel)
```

Heterogeneity across years (time effects):

```{r}
#| label: plotmeans-year

plotmeans(y ~ year, main = "Heterogeneity across years", data = Panel)
```

If these plots show big differences across countries or years, ignoring the panel structure will bias our estimates.

## Pooled OLS

The simplest approach is to ignore the panel structure and run ordinary OLS, treating every country-year as an independent observation:

```{r}
#| label: pooled-ols

ols <- lm(y ~ x1, data = Panel)
summary(ols)
```

```{r}
#| label: pooled-plot

plot(Panel$x1, Panel$y, pch = 19, xlab = "x1", ylab = "y")
abline(lm(Panel$y ~ Panel$x1), lwd = 3, col = "red")
```

This gives us one slope and one intercept. The implicit assumption is that *all countries are identical* --- there's no unobserved country-level factor affecting the outcome. That's a strong assumption.

## Fixed effects: the intuition

Fixed effects deals with heterogeneity by comparing each country **to itself** over time. The simplest way to understand this is to decompose each variable into two parts:

- **Between variation** = each country's average (how countries differ from each other)
- **Within variation** = each country's deviations from its own average (how a country changes over time)

Fixed effects throws away the between variation and uses only the within variation.

```{r}
#| label: fe-manual

# Decompose x1 into within and between components
btw.x <- ave(Panel$x1, Panel$country)   # Between: country means
wi.x  <- Panel$x1 - btw.x              # Within: deviations from country means

# Fixed effects = regression on within variation only
fixed.hand <- lm(Panel$y ~ wi.x)
summary(fixed.hand)
```

We can also include *both* components to see how they differ:

```{r}
#| label: fe-within-between

fixed.hand2 <- lm(Panel$y ~ wi.x + btw.x)
summary(fixed.hand2)
```

If the within and between coefficients are very different, it suggests that unobserved country-level confounders are biasing the between estimate --- exactly the situation where fixed effects helps.

## Fixed effects with dummy variables (LSDV)

An equivalent approach is to include a separate intercept for each country. This is the **Least Squares Dummy Variable** estimator:

```{r}
#| label: fe-lsdv

fixed.dum <- lm(y ~ x1 + factor(country) - 1, data = Panel)
summary(fixed.dum)
```

Each country gets its own intercept, but they all share the same slope on `x1`. We can visualize the fitted values:

```{r}
#| label: fe-lsdv-plot

yhat2 <- fixed.dum$fitted
scatterplot(yhat2 ~ Panel$x1 | Panel$country, boxplots = FALSE,
            xlab = "x1", ylab = "Fitted values", smooth = FALSE)
```

## Fixed effects with plm

In practice, we use the `plm` package, which handles the demeaning automatically and produces clean output:

```{r}
#| label: fe-plm

fixed <- plm(y ~ x1, data = Panel,
             index = c("country", "year"), model = "within")
summary(fixed)
```

We can extract the country-specific intercepts (the "fixed effects" themselves):

```{r}
#| label: fixef

fixef(fixed)
```

And formally test whether fixed effects are needed versus pooled OLS. The null hypothesis is that all individual intercepts are equal (i.e., pooled OLS is adequate):

```{r}
#| label: f-test-fe

pFtest(fixed, ols)
```

A significant result means the country effects are real and we should not ignore them.

## Random effects

The **random effects** model also accounts for individual heterogeneity, but assumes it is *random* and *uncorrelated with the predictors*. If that assumption holds, random effects is more efficient because it uses both within- and between-unit variation:

```{r}
#| label: random-effects

random <- plm(y ~ x1, data = Panel,
              index = c("country", "year"), model = "random")
summary(random)
```

## Comparing all three models

```{r}
#| label: compare-models
#| results: asis

stargazer(ols, fixed, random, type = "html",
          column.labels = c("Pooled OLS", "Fixed Effects", "Random Effects"))
```

## Hausman test: choosing between FE and RE

The **Hausman test** checks whether the random effects assumption (no correlation between individual effects and predictors) is valid.

- H0: Random effects is consistent (no correlation).
- H1: Only fixed effects is consistent.

```{r}
#| label: hausman

phtest(fixed, random)
```

If $p < 0.05$, reject H0 and use fixed effects. Otherwise, random effects is acceptable.

## Additional diagnostic tests

### Time fixed effects

Should we also control for year-specific shocks? We add year dummies and test whether they're jointly significant:

```{r}
#| label: time-fe

fixed.time <- plm(y ~ x1 + factor(year), data = Panel,
                  index = c("country", "year"), model = "within")
pFtest(fixed.time, fixed)
```

A significant result means time effects matter and should be included.

### Serial correlation

Panel data often exhibits serial correlation --- observations for the same unit in adjacent years are correlated. The Breusch-Godfrey test checks for this:

```{r}
#| label: serial-corr

pbgtest(fixed)
```

If significant, we may need to cluster our standard errors or use other corrections.

---

# Part 2: Two-Stage Models

Panel data addresses unobserved heterogeneity. But what about **reverse causality** or **omitted pathway** problems? This section covers three strategies: mediation analysis, sample selection correction, and instrumental variables.

We'll use the Acemoglu, Johnson, and Robinson (AJR) dataset, which studies how colonial-era settler mortality affected modern economic outcomes through the quality of institutions.

```{r}
#| label: load-ajr

library(haven)
library(mediation)
library(sampleSelection)
library(AER)

ajr2 <- read_dta("ajr2.dta")
```

## Mediation analysis

**Research question:** Does settler mortality affect GDP *through* the quality of institutions, or does it have a direct effect as well?

The causal chain we hypothesize is:

$$\text{Settler Mortality} \longrightarrow \text{Institutions (risk)} \longrightarrow \text{GDP}$$

Mediation analysis decomposes the total effect into a **direct** path and an **indirect** path (through the mediator).

### Step 1: Total effect (without the mediator)

First, what is the overall relationship between settler mortality and GDP?

```{r}
#| label: total-effect

total_model <- lm(loggdp ~ logmort0, data = ajr2)
summary(total_model)
```

### Step 2: Effect of X on the mediator

Does settler mortality predict institutional quality?

```{r}
#| label: mediator-model

mediator_model <- lm(risk ~ logmort0, data = ajr2)
summary(mediator_model)
```

### Step 3: Direct and indirect effects

When we include *both* settler mortality and institutions in the same model, we can see the direct effect of mortality (controlling for institutions) and use the coefficients to calculate the indirect effect:

```{r}
#| label: outcome-model

outcome_model <- lm(loggdp ~ risk + logmort0, data = ajr2)
summary(outcome_model)
```

The indirect effect is the product of the path from X to the mediator and the path from the mediator to Y:

```{r}
#| label: indirect-manual

coef_risk     <- coef(outcome_model)["risk"]
coef_logmort0 <- coef(mediator_model)["logmort0"]
indirect_effect <- coef_risk * coef_logmort0
indirect_effect
```

### Formal mediation test

The `mediate()` function uses bootstrapping to get proper confidence intervals for the indirect, direct, and total effects:

```{r}
#| label: mediation-formal

set.seed(12345)
mediation_result <- mediate(mediator_model, outcome_model,
                            treat = "logmort0", mediator = "risk",
                            sims = 1000, boot = TRUE)
summary(mediation_result)
```

```{r}
#| label: mediation-plot

plot(mediation_result)
```

**How to read the output:**

- **ACME** (Average Causal Mediation Effect) = the indirect effect through institutions
- **ADE** (Average Direct Effect) = the part of settler mortality's effect that doesn't go through institutions
- **Total Effect** = ACME + ADE
- **Proportion Mediated** = what fraction of the total effect operates through the mediator

## Sample selection models (Heckman correction)

Sometimes we observe the outcome only for a non-random subset of the population. For example, we only observe wages for people who *choose to work*. If the decision to work is correlated with wage determinants, naive OLS on the working subsample is biased.

The **Heckman selection model** corrects for this by jointly modeling the selection equation (who works?) and the outcome equation (what wage do workers earn?).

```{r}
#| label: heckman-setup

data("Mroz87", package = "sampleSelection")
```

First, the **naive approach** --- just running OLS on people who work:

```{r}
#| label: naive-ols

working_women <- subset(Mroz87, lfp == 1)
naive_model <- lm(log(wage) ~ educ + exper + I(exper^2) + age,
                  data = working_women)
summary(naive_model)
```

Now the **Heckman two-step estimator**. The selection equation includes variables that affect the decision to work (including `kids5` and `kids618`, which plausibly affect participation but not wages directly). The outcome equation models wages conditional on working:

```{r}
#| label: heckman-model

selection_formula <- lfp ~ nwifeinc + educ + exper + I(exper^2) + age + kids5 + kids618
outcome_formula   <- log(wage) ~ educ + exper + I(exper^2) + age

heckman_result <- selection(selection_formula, outcome_formula,
                            data = Mroz87, method = "2step")
summary(heckman_result)
```

Comparing the naive OLS with the selection-corrected estimates:

```{r}
#| label: heckman-compare
#| results: asis

stargazer(naive_model, heckman_result, type = "html",
          column.labels = c("Naive OLS", "Heckman Selection"))
```

If the **inverse Mills ratio** (the selection correction term, often labeled $\lambda$ or `rho`) is significant, it means selection bias was present and the correction matters.

## Instrumental variables

**Problem:** Institutions and GDP are likely jointly determined --- rich countries can afford better institutions, and better institutions produce growth. OLS on `GDP ~ institutions` conflates both directions.

**Solution:** Use **settler mortality** as an instrument for institutional quality. The logic: places where colonial settlers died at high rates got extractive institutions (because settlers didn't stay), and those institutions persisted. Settler mortality affects GDP *only through* its effect on institutions, not directly.

### First stage: Does the instrument predict the endogenous variable?

```{r}
#| label: first-stage

first_stage <- lm(risk ~ logmort0, data = ajr2)
summary(first_stage)
```

### Reduced form: Does the instrument affect the outcome?

```{r}
#| label: reduced-form

reduced_form <- lm(loggdp ~ logmort0, data = ajr2)
summary(reduced_form)
```

### IV estimation

The `ivreg()` function estimates the two-stage least squares model. The syntax uses `|` to separate the structural equation (left) from the instruments (right):

```{r}
#| label: iv-estimate

iv_model <- ivreg(loggdp ~ risk | logmort0, data = ajr2)
summary(iv_model)
```

### Comparing OLS vs. IV

```{r}
#| label: ols-iv-compare
#| results: asis

ols_biased <- lm(loggdp ~ risk, data = ajr2)

stargazer(ols_biased, iv_model, type = "html",
          column.labels = c("OLS (biased?)", "IV (consistent)"))
```

If the IV coefficient is substantially different from OLS, it suggests endogeneity was biasing the OLS estimate.

---

## Summary

| Problem | Method | Key Idea |
|---------|--------|----------|
| Unobserved heterogeneity | Panel data (FE/RE) | Compare units to themselves over time |
| Causal pathways | Mediation analysis | Decompose total effect into direct + indirect |
| Non-random samples | Heckman selection | Jointly model selection and outcome |
| Endogeneity / reverse causality | Instrumental variables | Isolate exogenous variation in predictor |
